---
title: "Week 11"
subtitle: "-"
---


## Types of spatial data
[@moragaSpatialStatisticsData2024]



- Data acquisition via google maps and open street maps

- Descriptive mapping (Open street maps, so on)

- Looking at distances (projections) and access to places 

- Isochronic maps, Voronoi maos



## Packages for downloading spatial data

https://www.paulamoraga.com/book-spatial/r-packages-to-download-open-spatial-data.html

### osmdata
[@moragaRspatialdataCollectionData2022]

```{r}
library(needs)
needs(osmdata, sf, tidyverse)
library(osmdata)
library(sf)
library(tidyverse)

# Define bounding box for Leipzig
bb_leipzig <- getbb("Leipzig", format_out = "polygon")

# Query OSM data for administrative boundaries of Leipzig
leipzig_admin <- opq(bb_leipzig) %>% 
  add_osm_feature(key = "boundary", value = "administrative") %>% 
  osmdata_sf() %>% 
  trim_osmdata(bb_leipzig)

# Query OSM data for kiosks in Leipzig
kiosks_leipzig <- opq(bb_leipzig) %>% 
  add_osm_feature(key = "shop", value = "bakery") %>% 
  osmdata_sf() %>% 
  trim_osmdata(bb_leipzig)

# Plotting using ggplot
ggplot() +
  geom_sf(data = leipzig_admin$osm_multipolygons) +
  geom_sf(data = kiosks_leipzig$osm_points) +
  theme_void()

```



## Voronoi-Zellen

Klappt noch nicht!

```{r}

# Make sure we have points
kiosk_pts <- kiosks_leipzig$osm_points

# 3. Transform to projected CRS (UTM Zone 32N = EPSG:25832)
crs_projected <- 25832
kiosk_pts_proj <- st_transform(kiosk_pts, crs = crs_projected)
leipzig_admin_proj <- st_transform(leipzig_admin$osm_multipolygons, crs = crs_projected)

# 4. Generate Voronoi polygons
voronoi_geom <- st_voronoi(st_union(kiosk_pts_proj))
voronoi_sf <- st_collection_extract(voronoi_geom) %>% st_sf()

# 5. Intersect with Leipzig boundary to clip the cells
voronoi_clipped <- st_intersection(voronoi_sf, leipzig_admin_proj)

# Add ID for coloring
voronoi_clipped <- voronoi_clipped %>% 
  mutate(cell_id = row_number())

# Plot
ggplot() +
  geom_sf(data = leipzig_admin_proj, fill = "white", color = "black") +
  geom_sf(data = voronoi_clipped, aes(fill = as.factor(cell_id)), alpha = 0.5, color = "black") +
  geom_sf(data = kiosk_pts_proj, color = "red", size = 1) +
  theme_void() +
  theme(legend.position = "none")

```




```{r}
# 1. Extract kiosk points and Leipzig polygon
kiosk_pts <- kiosks_leipzig$osm_points
leipzig_sf <- leipzig_admin$osm_multipolygons

# 2. Transform to projected CRS for better Voronoi accuracy
crs_projected <- 25832  # UTM Zone 32N for Leipzig area
kiosk_pts_proj <- st_transform(kiosk_pts, crs_projected)
leipzig_sf_proj <- st_transform(leipzig_sf, crs_projected)

envelope <- leipzig_sf_proj %>% 
  st_union() %>% 
  st_geometry()

kiosk_voronoi <- kiosk_pts_proj %>%
  st_geometry() %>%
  st_union() %>%
  st_voronoi(envelope = envelope) %>%
  st_collection_extract("POLYGON") %>%
  st_sf()

# 4. Assign CRS to Voronoi
st_crs(kiosk_voronoi) <- crs_projected

# 5. Intersect Voronoi cells with Leipzig boundary
#    and spatially join kiosk attributes
kiosk_voronoi_clipped <- kiosk_voronoi %>%
  st_intersection(leipzig_sf_proj) %>%
  st_join(kiosk_pts_proj, join = st_contains_properly)

kiosk_voronoi_clipped <- kiosk_voronoi_clipped %>%
  mutate(kiosk_id = row_number())

# 6. Plot the result, colored by kiosk name (or ID)
ggplot(kiosk_voronoi_clipped) +
  geom_sf(aes(fill = kiosk_id), color = "black") +
  theme_void()
```

# Isochronic maps (fromrome2rio)


Übung im Kurs: Leipziger Partnerstädte auf Weltkarte anzeigen


https://stackoverflow.com/questions/74198735/constructing-a-voronoi-diagram-in-r-with-data-from-openstreetmap